\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-05-04 main.tex
a good draft of the first part

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others
(increasing their payoff) at the expense of decreasing one's own payoff? During
the past decade, the literature on mechanisms that allow for cooperation, even
though it's seen to be at odd has been fruitful. One such mechanism is
repetition; the so called direct reciprocity. Theoretical researchers have been
using evolutionary models to understand how direct reciprocity allows
cooperation to evolve and which strategies are important for sustaining
cooperation.

Evolutionary game theory does not require individuals to be rational, instead
they adapt strategies based on mutation and exploration. Strategies with
high fitness are more likely to spread, either because the individuals who
adopt them have more offspring (fitness-based processes), or they are imitated
more often (pairwise comparison processes)~\cite{Wu2015}. The fitness of strategies,
and subsequently of the individuals, is not constant. Instead it depends on the
composition of the population. Individuals interact with other members of the
population according to their strategies. Their yielded payoffs, according to
pre defined games, are translated into fitness.

It is commonly assumed that fitness assimilates to the \textit{expected payoff},
which the mean payoff an individual achieved over the different types in the
population after multiple intercalations. Expected payoffs imply that
individuals have a perfect a memory. In order to estimate expected payoffs
individuals must be able to recall several of their interactions with each
player in the population. However, when modeling how they make decisions in each
round they assumed to have very limited memory. To be precise, most work on this
models focuses on naive subjects who can only choose from a restricted set of
strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016}, with a few notable
exceptions~\cite{Hauert1997, Stewart2016}.

This creates a curious inconsistency, which raises the following question: how
robust is our understanding of cooperation? We are not the first to question the
assumptions of models when estimating fitness. In~\cite{Roca2006} relax the
assumption that selection occurs much more slowly than the interaction between
individuals. They results shows that rapid selection affects evolutionary
dynamics in such a dramatic way that for some games it even changes the
stability of equilibria.

We consider two cases. Initially, we start by considering two extreme scenarios.
The first is the classical scenario of the expected payoffs and the alternative
scenario where individuals update their strategies only based on the very last
payoff they obtained. We refer to these as the \textit{stochastic} payoffs of
individuals. We do this for the donation games but also the important cases of
symmetric \(2 \times 2\) games. These include the prisoner's dilemma, the
snowdrift game, the stag hunt game and the harmony game.

In the later sections we allow individuals to use more memory. More specifically,
individuals update their strategies by considering up to the last two rounds,
whilst interacting with up to two different players.

@@

In order to account for the difference in the robustness of cooperation among
individuals based on their payoff memory, we consider a population of \(N\)
players, where \(N\) is even, and where mutations are sufficiently rare. At any
point in time the there are at most two different strategies are present in the
population. A \textit{resident} strategy and a \textit{mutant} strategy. Suppose
there are \(N - k\) players who use the resident strategy whereas \(k\) players
use the mutant strategy. Each step of the evolutionary process consists of two
stages, a game stage and an updating stage.

In the game stage, each player is randomly matched with some other player in
the population to interact in a number of turns.  The number of turns is
not fixed, on the contrast we consider that another interaction has a
probability $\delta$ to continue following each turn. At each turn the players can
choose to either cooperate (C) or to defect (D). The payoffs in a given round
depend on both player's decisions and are given by:

\begin{equation*}\label{eq:game}
  U = \begin{pmatrix} R & S  \\ T & P \end{pmatrix}
\end{equation*}

We assume herein that
individuals at most make use of simple \textit{reactive strategies} make
decisions in each round. Reactive strategy are a set of memory-one strategies
that only take into account the previous action of the opponent. Reactive
strategies can be written explicitly as a vector \(\in \R_{3}\). More
specifically a reactive strategy \(s\) is given by \(s=(y, p, q)\) where \(y\)
is the probability that the strategy opens with a cooperation and \(p, q\) are
the probabilities that the strategy cooperates given that the opponent
cooperated and defected equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, we
assume the learner adopts the role model's strategy based on the Fermi
distribution function. The relative influence of the payoffs on the
adopt the strategy of the other is controlled by an external parameter. The so
called intensity of selection,  $\beta\!\ge\!0$.

This basic evolutionary step is repeated until either the mutant strategy goes
extinct, or until it fixes in the population, in which case the mutant strategy
becomes the new resident strategy. After either outcome, we introduce a new
mutant strategy, uniformly chosen from all reactive strategies at random, and we
set the number of mutants to $k\!=\!1$. This process of mutation and
fixation/extinction is then iterated many times.

In this work we explore the effect on the payoffs in the updating stage. The
details of the process described in this section can be found in
appendix~\ref{appendix:methods}.

\section{Analysis of Stochastic Payoffs in the Donation game}\label{section:donation}

We first explore the effect of stochastic payoffs on the cooperative behaviour
by considering the extreme case that an individual's fitness is based on their
last interaction with one other individual. We compare this to the expected
payoffs. In this section we consider the donation game.
Each player can cooperate by providing a benefit \(b\) to the other
player at their cost \(c\), with \(0 < c < b\). Then, \(T=b, R=b-c, S=-c, P=0\),
and matrix~(\ref{eq:game}) is give by:

\begin{equation}\label{eq:donation_normal_form}
    \begin{pmatrix}
        b-c & -c  \\
        b & 0
    \end{pmatrix}
\end{equation}

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$. Assuming that the discount
factor~$\delta$ is comparably high, the opening move \(y\) is a transient effect
and has not effect on the outcome.
The left panels corresponds to the standard scenario considered in the previous
literature. It considers players who use expected payoffs to update their
strategies. The right panel shows the scenario considered herein, in which
players update their strategies based on their last round's payoff. The top
panels assume a benefit of 3, whereas the bottom assume that the benefit is 10.

The figure suggests that when updating is based on expected payoffs, players
tend to be more generous. The $q$-values are higher on average which suggests
that individuals tend to cooperating more after they are at the receiving end
of a defection. In addition, for both cases of \(b\), individuals tend to
be more cooperative. The average cooperation rate is strictly higher for expected
payoffs. This difference is statistically important, abd the effect is even more
obvious when the benefit is higher. More specifically, for \(b=10\) the average
cooperation rate drops from 97\% to 51\%.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}\vspace{-3cm}
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    We have run two simulations of the evolutionary process described in
    Section~\ref{section:methods} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. Also the resulting cooperation rate is smaller. On
    average, players cooperate roughly in half of all rounds. Parameters:
    $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

Figure~\ref{fig:cooperation_rate_over_benefit} further explores the effect of the
benefit on the cooperation rate. In all cases, the stochastic payoffs evolution
tends to reduce the evolving cooperation rate. For the stochastic payoffs,
benefit appears to not make a difference once \(b > 5\), and on average
the evolving cooperation rate is at 50\%. The effect of benefit on the expected
payoffs smooths out after \(b > 6\), and the expected payoffs estimating
the evolving rate to be at 90\%.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \vspace{.3cm}
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different benefit values.} 
Here, we vary the benefit of defection $b$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. Unless explicitly
varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$,
$\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$
time steps for each parameter
combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

Figure~\ref{fig:cooperation_rate_over_betas} illustrates the results for
different runs of the evolutionary process where we vary the strength of
selection. In the case of a very small \(\beta < \beta=10 ^ -1\) he process is
almost random, as the effect of the payoffs is very small. Once payoff begin to
matter the difference is once again evident, with the expected payoffs always
overestimating the evolved cooperation rate.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.} 
Here, we vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. Unless explicitly
varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$,
$\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$
time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

\section{Expected and stochastic payoffs in $2 \times 2$ games}\label{section:2_by_2_games}

>>>>

One of the most important applications of evolutionary game theory is the
evolution of cooperation. Why is it that some individuals choose to help others,
increasing their payoff, at the expense of decreasing one's own payoff? In
evolutionary game theory individuals are not required to be rational, instead
they adapt strategies based on mutation and exploration. Strategies are more
likely to spread if they have a high fitness either because the individuals who
adopt them have more offsprings, or because they are imitated more often. The
fitness of a strategy is not constant but depends on the composition of the
population. Individuals interact based on their strategies with other
members of the population and the payoffs they yield are translated into
fitness.

It is commonly assumed that fitness is equivalent to the mean payoff an
individual achieves over the different types in the population, over multiple
interactions. These payoffs assume that individuals can interact with the entire
population several times and remember each and every outcome. Thus, they imply
that individuals have a perfect memory.
However, in their pairwise interactions when they make decisions in each turn
they are assumed to have very limited memory. To be precise most of
the works in the literature,  with a few notable exceptions~\cite{Hauert1997,
Stewart2016}, focus on naive subjects who can only choose from a restricted set
of strategies~\cite{Nowak1992tit}, or who do not remember anything beyond the
outcome of the very last round~\cite{Baek2016}.

The perfect memory assumption is not only unrealistic but it also creates this
curious inconsistency. This has lead us to question how robust is our
understanding of cooperation. In this work we explore whether direct
reciprocity can evolve if individuals only remember a minimum of social
information. Though we are not the first to question the assumptions of
estimating fitness~\cite{Roca2006}, we are the first to explore the effect
of payoff memory.

Initially, we consider two extreme scenarios. The first is the classical
scenario of the expected payoffs and the alternative scenario where individuals
update their strategies only based on the very last payoff they obtained. In the
later sections, we allow individuals more memory. More specifically, they can
remember up to two turns and up to two interactions.
We present results on  several social dilemmas which include the prisoner's
dilemma and the donation game, the snowdrift game, the stag hunt game and the
harmony game.

The remainder of the paper is organized as follows. Section~\ref{section:model}
describes the model. Section~\ref{section:results} presents the results of the
simulations. Finally, section~\ref{section:conclusions} outlines the main
conclusions.

@@

We consider a population of \(N\) players (\(N\) is even) where mutations are
sufficiently rare. Thus, at any point in time there are at most two
different strategies present in the population; a \textit{resident} strategy
and a \textit{mutant} strategy. We assume a pairwise process where
strategies spread because they are imitated more often.

Each step of the evolutionary process consists of two stages, a game stage and
an updating stage. In the game stage each individual is randomly matched with
some other individual in the population to interact for a number of turns, where
subsequent turns occur with a fixed probability $\delta$. At each turn they
can choose to either cooperate (\(C\)) or to defect (\(D\)), and thus, at
each turn the possible outcomes are \(CC, CD, DC\) and \(DD\). The payoffs
depend on the outcome. If both cooperate they receive the reward payoff
\(R\), whereas if both defect they receive the punishment payoff \(P\).
If one cooperates but the other defects, the defector receives the temptation to
defect, \(T\), whereas the cooperator receives the sucker's payoff, \(S\).
We denote the payoffs of an individual as \(\mathcal{U} = (R, S, T, P)\).

We assume herein that individuals use \textit{reactive strategies} to make
decisions in each turn. Reactive strategies are a set of memory-one strategies
that only take into account the previous action of the opponent. They can be
written explicitly as a vector \(\in \R_{3}\), more specifically, a reactive
strategy \(s\) is given by \(s=(y, p, q)\) where \(y\) is the probability that
the strategy opens with a cooperation and \(p, q\) are the probabilities that
the strategy cooperates given that the opponent cooperated and defected
equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given that the learner's payoff $u_L\!\in\!
\mathcal{U}$ and that the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, we
assume the learner adopts the role model's strategy based on the Fermi
distribution function,

where $\beta\!\ge\!0$ is the relative influence of the payoffs on adopting the 
strategy of the other. We refer to $\beta$ as the intensity of selection.

This basic evolutionary step is repeated until either the mutant strategy goes
extinct, or until it fixes in the population. If the mutant fixes in the
population then the mutant strategy becomes the new resident strategy. After
either outcome we introduce a new mutant strategy uniformly chosen from all
reactive strategies at random, and we set the number of mutants to $1$. This
process of mutation and fixation/extinction is then iterated many times.

The perfect memory assumption occurs at the updating stage. The learner and the
role model are assumed to interact with a representative sample of the
population, and they remember all interactions they participate in. Thus, they
are updating payoffs are based on the mean payoff they achieved over all the
interactions. These payoffs are refer to as the expected payoffs. We will
compare the expected payoffs to payoffs that are calculated when the role model
and learner do not remember all their interactions. In order to account for the
effect of these different methods we explore the cooperation rate within the
resident population over multiple generations. More details on our methodology
are found in Appendix~\ref{appendix:methods}.

\section{Results}\label{section:results}
\subsection{Updating payoffs based on the last round with another member of the population}\label{section:donation}

In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population. We compare
this to the expected payoffs. We assume that individuals interact in a donation
game  where each can cooperate by providing a benefit \(b\) to the other
player at their cost \(c\), with \(0 < c < b\). Thus, \(T=b, R=b-c, S=-c, P=0\).

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus the opening move \(y\) is a transient effect and has
not effect on the outcome. The left panels corresponds to the standard scenario
considered in the literature. It considers players who use expected payoffs to
update their strategies. The right panel shows the scenario considered herein,
in which players update their strategies based on their last round's payoff. The
top panels assume a benefit \(b\) of 3, whereas the bottom assume a benefit of
10.

The figure suggests that when updating is based on expected payoffs, players
tend to be more generous and more cooperative. The $q$-values are higher on
average which suggests that individuals tend to be more generous when it comes
to cooperating after being at the receiving end of a defection. %This is a known result no? Nowak 1992
The average cooperation rate within the resident population is strictly higher
when expected payoffs are considered. The difference between the two methods for
both the values of \(b\) are statistically significant. This contrast becomes
more obvious for \(b=10\). More specifically the average cooperation drops from
97\% to 51\%. The residents of the population cooperated on average 97\% of the
times in one case and only 57\% in the other.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}\vspace{-3cm}
    \includegraphics[width=.75\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and last round with one interaction payoffs.} 
    We have run two simulations of the evolutionary process described in
    section~\ref{section:model} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. Parameters:
    $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

We further explore the effect of the benefit,
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that the
expected payoffs always overestimate cooperation. For the limited memory payoffs
the cooperation rate remains unchanged at approximately 50\% once \(b = 5\). In
specific simulations we have performed, if only the last round payoff is
considered then the highest average cooperation rate that was achieved was 0.51.
In comparison, when the expected payoffs were used the highest average
cooperation rate was 0.97.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
  \end{subfigure}
  \caption{{\bf The evolution of cooperation for different benefit values.} 
  We vary the benefit of defection $b$. In all cases, expected payoffs appear to
  overestimate the average cooperation rate the population achieves. ({\bf A})
  the probabilities \(p, q\) for resident population over \(10^7\) time steps
  for each benefit value. ({\bf B}) The cooperation rate within the resident population
  (averaged over all games and over all time steps) over the benefit.
  Unless
  explicitly varied, the parameters of the simulation are $N\!=\!100$,
  $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run
  for $T\!=\!5\times 10^6$ time steps for each parameter
  combination.}\label{fig:cooperation_rate_over_benefit}
\end{figure}

Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process where we vary the strength of selection. For
weak selection, \(\beta < 1\), the updating payoffs have no effect. The evolved
population and the average cooperating rate are the same for both approaches.
However for strong selection, it can be seen that the cooperating rate increases
with \(beta\), when expected payoffs are used. However, when limited memory
payoffs are used. the cooperating rate decreases.

\begin{figure}[!htbp]
  \centering
  \begin{subfigure}{.65\textwidth}
    \centering
    \includegraphics[width=\textwidth]{static/expected_for_selection_strenght.pdf}
    \includegraphics[width=\textwidth]{static/stochastic_for_selection_strenght.pdf}
  \end{subfigure}%
  \begin{subfigure}{.35\textwidth}
    \vspace{.3cm}
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
  \end{subfigure}
\caption{{\bf The evolution of cooperation for different selection strength values.} 
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\).
Unless explicitly varied, the parameters of the simulation are $N\!=\!100$,
$b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$T\!=\!5\times 10^6$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

\subsection{Expected and stochastic payoffs in $2 \times 2$ games}\label{section:2_by_2_games}

\section{Conclusions}\label{section:conclusions}
